---
title: "Estimate the migration rate between two areas from tag return data"
author: "Philippe Ziegler and Paul Burch"
date: "5 June 2017"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Estimate migration rates from tag returns}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: migrate.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

The vignette estimates the rate of fish migration between two areas from tag
returns using a catch conditioned modification of @Hilborn1990movement.

# Methods


The model simulated two areas 1 and 2, and followed an annual sequence 
of tag release with release mortality in area 1, then application of tag shedding, 
half of natural mortality and removal of recaptures by fishing in both areas, 
fish movement from area 1 to area 2, and finally application of the second half 
of natural mortality to the remaining numbers of tagged fish. 

We assumed that tagged and released fish $relT_{1,y}$ in area 1 at the start 
of year $y$ were subject to tag-release mortality ($\phi$) such that available 
tags $T_{1,y}$:  

$$T_{1,y}= relT_{1,y} * (1-\phi)$$

Each year, tag-shedding $s$ and half of annual natural mortality $M$ were 
applied first to all available tags in an area $r$: 

$$T_{r,y} = T_{r,y} * (1-\lambda)*exp(-0.5M)$$

before fishing removed recaptures $Rec$ from the population of tagged fish: 

$$T_{r,y} = T_{r,y} - Rec_{r,y} = T_{a,y}-\frac{T_{r,y}*C_{r,y}}{B_{r,y}}$$

where $C_{r,y}$ is the catch in numbers and $B_{r,y}$ is the vulnerable 
population size in numbers. After fishing, the number of tagged fish $T_move$ 
that moved from area 1 to area 2 was given by: 

$$T_{move} = move*T_{1,y}$$

where $move$ is the annual movement rate. The number of tagged fish in area 1 and 
area 2 were then updated: 

$$T_{1,y} = T_{1,y} - T_{move}$$

$$T_{2,y} = T_{2,y} - T_{move}$$

Finally, the second half of annual natural mortality $M$ was applied: 

$$T_{r,y} = T_{r,y}*exp(-0.5M)$$

Since fishing occurred in the simulations before movement, no tagged fish were 
recaptured within-season in area 2 from tags released in area 1. Annual cohorts 
of tag-releases were treated separately in the simulation. Movement rate was 
estimated by minimising the negative multinomial log likelihood of observed and 
expected numbers of recaptures in either area 1, area 2, or both. 

# Simulation

The model is demonstrated using simulated data

Load the **migrate** package

```{r}
library(migrate)
```


Specify the parameters used in the model

```{r}
yrs <- 2007:2015
relT <- rep( 1000, length(yrs))
# Annual Tag loss rate (for single-tag model) 
tagloss	<- rep(0.006, length(yrs))	
# Natural Mortality
M <- 0.155
# Release mortality
relM <- 0.1
# Proportion of fish that migrate
true_move	<- 0.01		
# Catch
C1 <- rep(10000, length(yrs))	
C2 <- rep(10000, length(yrs))
# Abundance in the two areas
B1 <- rep(1000000, length(yrs))	
B2 <- rep(1000000, length(yrs))
```

Define some matrices to store the tags available and the recaptures

```{r}
# T: Number of tags available in Areas 1 and 2 (after half of M)
T1 <- T2	<- matrix(0,nrow=length(yrs),ncol=length(yrs),dimnames=list(yrs,yrs))
# Number of tags being recaptured
R1 <- R2 <- T1
# Number of tags moving
T_move <- T1
```

Test: Apply tag releases, release M, half of M, remove & store recaptures, estimate and do movement

```{r}
for (x in 1:length(yrs)) {
  # Release tags
	T1[x,x]	<- relT[x]		
	# Apply release mortality
	T1[x,x]	<- (1-relM) * T1[x,x]		
	for (y in x:length(yrs)) {			
			# Available Tags after applying 0.5 M
			T1[x,y] <- T1[x,y] * exp(-0.5*M)
			T2[x,y] <- T2[x,y] * exp(-0.5*M)			 
			# Tagloss rate
			T1[x,y] <- T1[x,y] * exp(-tagloss[y])				 
			T2[x,y] <- T2[x,y] * exp(-tagloss[y])
			# Recaptures (by Petersen)
			R1[x,y] <- rbinom(n=1,size=round(T1[x,y],0),prob=C1[y]/B1[y])	
			R2[x,y]	<- rbinom(n=1,size=round(T2[x,y],0),prob=C2[y]/B2[y])	
			if((y+1) <= length(yrs)) {				# Prevent error
				T_move[x,y+1]	<- (T1[x,y] - R1[x,y])*true_move			# Movement after removals [= T*move*(1-move)**(y-x-1)]
				T1[x,y+1]	<- T1[x,y] - R1[x,y] - T_move[x,y+1]	# = T*(1-move)**(y-x)
				T2[x,y+1]	<- T2[x,y] - R2[x,y] + T_move[x,y+1]	# = 1-T*(1-move)**(y-x)
			}
			T1[x,y]		<- T1[x,y] * exp(-0.5*M)	# Apply 0.5 M - do not apply so T1&T2 remains mid-season numbers
			T2[x,y]		<- T2[x,y] * exp(-0.5*M)	# 
	}
}
```

Examine the matrices of tags for the two areas

```{r}
T1
T2
```

Examine the matrices of recaptures for the two areas

```{r}
R1
R2
```

Define the parameters

```{r}
obs <- list(yrs=yrs,relT=relT,R1=R1,R2=R2,prec1=C1/B1,
            prec2=C2/B2,M=M,relM=relM,tagloss=tagloss)
para <- logitT(c(move=0.011694))	# move
```

Now fit to the simulated data using the least squares method

```{r}
fit_ls <- move_sim(obs,para,rfit="R2",lltype="prop")
```

Print the results

```{r}
cbind(Est=ilogitT(fit_ls$coef),
	    Lwr=ilogitT(fit_ls$coef-2*fit_ls$coef.se),
	    Upr=ilogitT(fit_ls$coef+2*fit_ls$coef.se))
```

Now fit to the simulated data using the maximum likelihood method

```{r}
fit_ml <- move_sim(obs,para,rfit="R2",lltype="multinom")
```

Print the results

```{r}
cbind(Est=ilogitT(fit_ml$coef),
	    Lwr=ilogitT(fit_ml$coef-2*fit_ml$coef.se),
	    Upr=ilogitT(fit_ml$coef+2*fit_ml$coef.se))
```


# References


